---
title: "Quantifying Faceoff Value and Identifying High Leverage Performance in the NHL"
author:
  - Chloe Guagliano
  - Tess Hinkle
  - Jack Pallotta
date: "July 24, 2026"
format: 
  html:
    toc: true
    toc-location: left
    toc-title: "Contents"
    theme: cosmo
    html-math-method: katex
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

https://rentosaijo.github.io/nhlscraper/logo.png \<- nhlscraper logo if we want to put it in our report

## Introduction

In hockey, faceoffs are used to (re)start playing time. Faceoffs can occur after any stoppage, such as period-end, icing, goal, and penalties. Faceoffs can also occur at varying leverage states in a game. High leverage in hockey is when certain states of a game, such as score and strength, allow for a team and players to have a higher probability of success throughout the game.

## Data

The data used throughout the project is pulled from the publicly available [nhlscraper](https://rentosaijo.github.io/nhlscraper/) R package created by Rento Saijo. The data that we specifically looked at from the nhlscraper package was NHL game play by play data for the 2021-2022 to the 2025-2026 seasons. We ultimately worked with two unique data sets compiled from one large 5 season play by play data set. One of our data sets looks at the events that occur within 5 seconds of a faceoff. The other data set looks at the faceoff event itself and mirrors the faceoff for hte home and away team.

```{r}
require(ggplot2)
require(tidyverse)
require(pROC)
theme_set(theme_bw())
events_after_faceoff2 = readRDS("Shiny/events_after_faceoff2.rds")
events_model = readRDS("Shiny/events_model.rds")
train_data = readRDS("Shiny/train_data.rds")
nhl.mod = readRDS("Shiny/nhl_mod.RDS")
```

### EDA

#### Quantifying Faceoff Value

The following graphs are EDA for the Quantifying Faceoff Value question. The bar chart of shot attempts between three NHL teams (Carolina, Chicago, and Pittsburgh) depicts that the majority of shot attempts for teams occur in the offensive zone. Also, over 50% of shot attempts are successful for these teams.

```{r}
#| fig-cap: "The bar chart shows the count of shot attempts in each zone and whether the shot attempt was successful or not."
barplot4 = events_after_faceoff2 |>
  filter(eventOwnerTeamId == 12 | eventOwnerTeamId == 16 | eventOwnerTeamId == 5, !is.na(zoneCode)) |> 
  mutate(teams = case_when(eventOwnerTeamId == 12 ~ "CAR",
                           eventOwnerTeamId == 16 ~ "CHI",
                           eventOwnerTeamId == 5 ~ "PIT")) |>
  ggplot(aes(x = zoneCode, fill = as.factor(is_shot_atmpt))) +
  geom_bar() +
  facet_wrap(~ teams) +
  labs(fill = "Is Shot Attempt", title = "Shot Attempts between three NHL Teams") +
  theme(legend.position = "bottom")
barplot4
```

Describe the data you’re using in detail, where you accessed it, along with relevant exploratory data analysis (EDA). You should also include descriptions of any relevant data pre-processing steps (e.g., whether you consider specific observations, create any meaningful features, etc.---but don't mention minor steps like column type conversion, filtering out unnecessary rows)

## Methods

Describe the modeling techniques you chose, their assumptions, justifications for why they are appropriate for the problem, and how you're comparing/evaluating the different methods.

### High Leverage Performance

methods for High leverage performance

### Quantifying Faceoff Value
A simple logistic regression was first tried due to the binary variable of future unblocked shot attempt. However, when coordinates were looked at as a predictor, we decided to transition to a generalized additive model for smoothing. The generalized additive model took hours to process so the model was transitioned to a big additive model to cut down on processing time.

The data was split into a 75% training and a 25% testing Holdout method to check for accuracy and to help ensure fast processing speed with a relatively large data set unlike k-fold cross-validation which would take longer.


## Results

Describe your results. This can include tables and plots showing your results, as well as text describing how your models worked and the appropriate interpretations of the relevant output. (Note: Don’t just write out the textbook interpretations of all model coefficients. Instead, interpret the output that is relevant for your question of interest that is framed in the introduction)

### High Leverage Performance

### Quantifying Faceoff Value

The future shot probability model resulted in high accuracy at predicting when a future shot after a faceoff will occur. The following plot is a calibration plot of the model showing how the model accurately predicts on the low and high end of shot probabilities, but is missing predictions in the middle range.

```{r}
shot_results = train_data |>
  select(future_shot, eventTypeDescKey, xCoord, yCoord,distance, scoreState, strengthState,
         secondsElapsedInGame, angle, isEmptyNetFor, isEmptyNetAgainst) |>
  drop_na() |>
  mutate(shot_prob = predict(nhl.mod, type = "response"),
         pred_decile2 = ntile(shot_prob, 10))

#checks each level of predictions withe the actual results
shot_calibration_check =  shot_results |>
  group_by(pred_decile2) |>
  summarize(
    predicted = mean(shot_prob),
    actual = mean(future_shot),
    n = n(),
    .groups = "drop")
shot_calibration_check
ggplot(shot_calibration_check,
       aes(x = predicted, y = actual)) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed") +
  geom_line(color = "blue") +
  geom_point(shape = 15, size = 3, color = "blue", alpha = 0.5) +
  labs(x = "Predicted Probability", y = "Observed Probability",
       title = "Calibration Plot") +
  coord_equal() +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5,  size = 15))
```

The following plot is a ROC plot showing how the model is also quite accurate.

```{r}
roc_obj2 = roc(
  response = shot_results$future_shot,
  predictor = shot_results$shot_prob,
  quiet = TRUE)

auc(roc_obj2)

shot_roc = tibble(threshold = c(roc_obj2$thresholds),
                  specificity = roc_obj2$specificities,
                  sensitivity = roc_obj2$sensitivities)

shot_roc |> 
  ggplot(aes(x = 1 - specificity , y = sensitivity)) + #1-specificity (false pos. rate)
  geom_path() +
  geom_abline(slope = 1, intercept = 0, 
              linetype = "dashed") + 
  theme_bw() + 
  labs(title = "ROC Plot of Predicted Future Shot Attempts", y = "True Positive Rate", x = "False Positive Rate") +
  theme(plot.title = element_text(hjust = 0.5, size = 15))
```

## Discussion

Give your conclusions and summarize what you have learned with regards to your question of interest. Are there any limitations with the approaches you used? What do you think are the next steps to follow-up your project?

### High Leverage Performance

### Quantifying Faceoff Value
Ultimately we are able to determine if events directly following a faceoff will result in a future unblocked shot attempt. However, these probabilities are either on the very low end (< 0.2) or on the higher end (>0.60) most likely due to the use if o binary predictor.

## Appendix: A quick tutorial

**(Feel free to remove this section when you submit)**

This a Quarto document. To learn more about Quarto see <https://quarto.org>. You can use the Render button to see what it looks like in HTML.

### Text formatting

Text can be bolded with **double asterisks** and italicized with *single asterisks*. Monospace text, such as for short code snippets, uses `backticks`. (Note these are different from quotation marks or apostrophes.) Links are written [like this](http://example.com/).

Bulleted lists can be written with asterisks:

- Each item starts on a new line with an asterisk.
- Items should start on the beginning of the line.
- Leave blank lines after the end of the list so the list does not continue.

Mathematics can be written with LaTeX syntax using dollar signs. For instance, using single dollar signs we can write inline math: $(-b \pm \sqrt{b^2 - 4ac})/2a$.

To write math in "display style", i.e. displayed on its own line centered on the page, we use double dollar signs: $$x^2 + y^2 = 1$$

### Code blocks

Code blocks are evaluated sequentially when you hit Render. As the code runs, `R` prints out which block is running, so naming blocks is useful if you want to know which one takes a long time. After the block name, you can specify [chunk options](https://yihui.org/knitr/options/). For example, `echo` controls whether the code is printed in the document. By default, output is printed in the document in monospace:

```{r, echo = FALSE}
head(mtcars)
```

Chunk options can also be written inside the code block, which is helpful for really long options, as we'll see soon.

```{r}
#| echo: false
head(mtcars)
```

### Figures

If a code block produces a plot or figure, this figure will automatically be inserted inline in the report. That is, it will be inserted exactly where the code block is.

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-cap: "This is a caption. It should explain what's in the figure and what's interesting about it. For instance: There is a negative, strong linear correlation between miles per gallon and horsepower for US cars in the 1970s."

library(tidyverse)
mtcars |> 
  ggplot(aes(x = mpg, y = hp)) +
  geom_point() +
  labs(x = "Miles per gallon",
       y = "Horsepower")
```

Notice the use of `fig-width` and `fig-height` to control the figure's size (in inches). These control the sizes given to `R` when it generates the plot, so `R` proportionally adjusts the font sizes to be large enough.

### Tables

Use the `knitr::kable()` function to print tables as HTML:

```{r}
mtcars |> 
  slice(1:5) |> 
  knitr::kable()
```

We can summarize model results with a table. For instance, suppose we fit a linear regression model:

```{r}
#| echo: true
model1 <- lm(mpg ~ disp + hp + drat, data = mtcars)
```

It is *not* appropriate to simply print `summary(model1)` into the report. If we want the reader to understand what models we have fit and what their results are, we should provide a nicely formatted table. A simple option is to use the `tidy()` function from the `broom` package to get a data frame of the model fit, and simply report that as a table.

```{r }
#| results: "asis"
#| tbl-cap: "Predicting fuel economy using vehicle features."
library(broom)
model1 |> 
  tidy() |>
  knitr::kable(digits = 2,
               col.names = c("Term", "Estimate", "SE", "t", "p"))
```
