---
title: "Quantifying Faceoff Value and Identifying High Leverage Performance in the NHL"
author:
  - Chloe Guagliano
  - Tess Hinkle
  - Jack Pallotta
date: "July 24, 2026"
format: 
  html:
    toc: true
    toc-location: left
    toc-title: "Contents"
    theme: cosmo
    html-math-method: katex
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
embed-resources: true
editor: visual
---

## Introduction

Every hockey game includes about 55 to 60 faceoff events. There is an opening faceoff that takes place at center ice to mark the beginning of each period and additional, dynamically occurring faceoffs that take place at any one of nine faceoff dot locations after each stoppage in play. The faceoff event is a primary responsibility of the center position in hockey, but any player is eligible to take a faceoff besides the goaltender. While a fan may reduce faceoff events to mere breaks or pauses in the speed and intensity of gameplay, well-respected National Hockey League players and coaches attest that the outcome of faceoff events carry meaningful implications on hockey games and offer competitive advantages to the team who is able to win it. In a [morning skate interview](https://lakingsinsider.com/2014/01/30/january-30-morning-skate-quotes-voynov-kopitar-stoll/) with LA Kings Insider, two-time Stanley Cup Champion Jarret Stoll stated that winning the battle amongst elite centers begins with faceoffs, linking faceoff wins to increased puck possession and opportunities to generate scoring chances (). Since retiring from his playing career, Stoll has transitioned into a player development coach role with the Los Angeles Kings and continues to discuss the nuances of faceoff events in the public sphere [link](https://www.sportsnet.ca/hockey/nhl/stoll-breaks-kings-miracle-0-9-second-faceoff-play/).

When watching a hockey game, it is clear that certain game states are more determinant of winning the game than others. As a result, the relative importance of a faceoff win (or any event, for that matter) is not static; it varies based on the game state. For the purposes of this project, leverage can be defined as the power of a faceoff outcome to greatly impact the game outcome. In certain game states that are of higher leverage, winning or losing a given faceoff could greatly impact whether the faceoff team wins or loses the game. In lower leverage game states, winning or losing a given faceoff is less determinant of winning or losing the game.

The following two faceoff events found in the 2025-26 NHL season are a prime example of how leverage varies by game state: **insert gifs**

Although there is benefit to winning both of the above faceoffs and both winning teams are able to generate a shot attempt from the possession sequence following their faceoff win, the overall importance of that win and shot attempt on the game outcome vary significantly.

In Clip 1, Flames forward Tyson Gross wins a neutral zone faceoff midway through the 3rd period while defending a four-goal lead. This faceoff has low leverage, because the game stakes when the faceoff is being taken is low. Regardless of the outcome of the faceoff, Calgary likely has this game in-hand. Alternatively, Clip 2 features a faceoff between J.G. Pageau and Anze Kopitar in the dying seconds of a one-goal game. The Islanders have a six-on-four advantage, with a powerplay and the goalie pulled for an extra attacker in hopes of pushing this game to overtime. This faceoff has higher leverage, because the stakes are high.

While leverage can identify game states where faceoffs matter most, it does not consider the outcome of the faceoff. If Pageau wins the faceoff, the Islanders establish possession with an opportunity to even up the game. On the other hand, if Kopitar wins, the Kings are in a significantly better position to defend their late game lead and come away with the win. Each faceoff serves as a cross-road for the flow of the game. Subsequent scoring opportunities a team can generate varies greatly based on whether or not they win a given faceoff and gain possession of the puck.

Faceoff win percentage, calculated as a player’s number of faceoff wins divided by their total number of faceoffs taken, is the primary metric used to measure player faceoff contributions in public hockey circles. However, this metric treats every faceoff as if they were of equal importance. As a result, the greater contributions of players who consistently win high leverage faceoffs that are most meaningful to game outcomes are not being captured. Faceoff win percentage also only captures whether a given faceoff is won or lost, not accounting for events or scoring opportunities to follow.

This research aims to contextualize faceoff leverage through the use of leverage groups and quantify the value of a faceoff win through short-term measures of success and expected benefit. Expected goals (xG) within five seconds of a faceoff were selected as our measure of short-term success. A five-second post-faceoff window was selected for simplicity and due to the structure of the NHL play-by-play data.

## Data

```{r}
require(ggplot2)
require(tidyverse)
require(pROC)
require(gt)
theme_set(theme_bw())
events_after_faceoff2 = readRDS("Shiny/events_after_faceoff2.rds")
events_model = readRDS("Shiny/events_model.rds")
train_data = readRDS("Shiny/train_data.rds")
nhl.mod = readRDS("Shiny/nhl_mod.RDS")
leverageVariables = readRDS("leverageVariables.rds")
```

The data used throughout the project is pulled from the publicly available [nhlscraper](https://rentosaijo.github.io/nhlscraper/) R package created by Rento Saijo. We specifically looked at NHL play-by-play data for the 2021-2022 to 2025-2026 seasons due to improvements in game scoring and shot tracking. Since the NHL play-by-play consists of discrete event data rather than continuous puck and player tracking, the exact duration of influence for a given faceoff can’t be accurately observed. A five-second window provides a practical approximation of the post-faceoff sequence while limiting unrelated downstream events.

Two unique data sets were compiled from one large five-season play-by-play data set. The raw NHL play-by-play data provides information regarding a single faceoff event from the perspective of the winning team. Our first dataset is a restructured version of this, specifically for faceoff events, so that each faceoff event has two corresponding rows of data for each team involved. The mirroring of the faceoff dataset enabled us to explore faceoff events from each team’s perspective, which was very beneficial for modeling. Second, another dataset focuses on events that occur within five seconds of a faceoff with a binary unblocked shot attempt variable for the interactive Shiny app.Allowing the events occurring after the facoff to have their own unique rows allows for the Shiny app to enable multiple event sequences instead of just looking at a faceoff.

```{r}
#first data set 
```

```{r}
events_model |>
  select(gameId, periodNumber, faceoff_time, secondsElapsedInGame, eventTypeDescKey,
         xCoord, yCoord, strengthState, scoreState, future_shot) |>
  head() |>
  gt()

```

Post-faceoff windows without a continuous five-second window of play to generate a scoring opportunity remain part of the dataset. These “censored” values arise when a penalty is committed or the puck is flipped out of play. An available scoring opportunity variable is calculated in seconds as a feature during the modeling phase to evaluate scoring opportunity over the subsequent five seconds of a faceoff.

### EDA

Before creating any models, exploring the play-by-play data helped reveal which features to consider when evaluating the leverage and expected value of a faceoff win.

#### High Leverage Performance

The first component in estimating leverage is the win probability for the two teams facing off before the puck drop. The value of a goal is not constant, as some goals can alter the final game outcome, while others have minimal effect in a blowout scenario.

A plot of the relationship between time remaining and empirical win percentage, faceted by the goal differential, shows that time effects are nonlinear with empirical win percentage. Once a multi-goal lead is built, empirical win probability tails off as the game nears its conclusion. For score differentials greater than three goals, observations were grouped into a ±3 differential bucket for interpretability.

Strength state, or the on-ice manpower situation between opposing teams, is another factor to consider when quantifying leverage. A plot of empirical win percentage by strength state indicates that win probability changes given the man differential. The data contained a limited number of ±3 manpower situations, so ±2 with an empty net for or against were grouped to create the "EN ±2" strength state.

Considering only a five-second window within faceoffs, distance from the attacking net plays a factor in the expected value of winning the draw. Winning an offensive zone faceoff produces a much higher mean xG within five seconds of a faceoff.

First Plot

```{r}
leverageVariables |>
  distinct(eventId, .keep_all = TRUE) |>
  filter(isOT == "Regulation") |>
  mutate(timeBin = cut(secondsRemaining, breaks = seq(0, 3600, by = 300), include.lowest = TRUE)) |>
  group_by(goalDifferential, timeBin) |>
  summarize(secondsRemaining = mean(secondsRemaining),
            winPct = mean(wonGame),
            .groups = "drop") |>
  ggplot(aes(secondsRemaining, winPct,
             color = goalDifferential)) +
  geom_line(linewidth = 1) +
  scale_x_reverse() +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.25)) +
  scale_color_manual(
  values = c("-3" = "#B2182B", "-2" = "#EF8A62", "-1" = "#FDB863",
             "0" = "#404040", "1" = "#92C5DE", "2" = "#67A9CF", "3" = "#2166AC")) +
  labs(title = "Empirical Win Percentage by Goal Differential and Time Remaining",
       x = "Seconds Remaining",
       y = "Empirical Win Percentage",
       color = "Score Differential") +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
        axis.title = element_text(size = 12))
```

Second Plot

```{r}
leverageVariables |>
  distinct(eventId, .keep_all = TRUE) |>
  group_by(situationCode) |>
  summarize(WinPct = mean(wonGame), Faceoffs = n()) |>
  ggplot(aes(x = reorder(situationCode, WinPct), y = WinPct, fill = WinPct)) +
  geom_col() +
  scale_fill_gradient2(low = "red", mid = "gold", high = "forestgreen",
                       midpoint = 0.5, labels = scales::percent) +
  geom_text(aes(label = scales::percent(WinPct, accuracy = 0.1)),
            size = 3, vjust = -0.5) +
  labs(title = "Empirical Win Percentage by Situation Code",
       x = "Situation Code",
       y = "Empirical Win Percentage",
       fill = "Win Percentage") +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
        axis.title = element_text(size = 12))
```

Third Plot

```{r}
leverageVariables |>
  distinct(eventId, .keep_all = TRUE) |>
  group_by(zoneCode, faceoffWon) |>
  summarize(mean_xG = mean(xGFor5), .groups = "drop") |>
  mutate(faceoffWon = factor(faceoffWon,
                             levels = c(0, 1),
                             labels = c("Loss", "Win"))) |>
  ggplot(aes(zoneCode, mean_xG, fill = faceoffWon)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  labs(title = "Mean xG Within 5 Seconds by Faceoff Zone",
       x = "Faceoff Zone",
       y = "Mean xG Within 5 Seconds",
       fill = "Faceoff Outcome") +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
        axis.title = element_text(size = 12))
```

#### Quantifying Faceoff Value

The following graphs are EDA for the Quantifying Faceoff Value question. The bar chart of shot attempts between three NHL teams (Carolina, Chicago, and Pittsburgh) depicts that the majority of shot attempts for teams occur in the offensive zone. Also, over 50% of shot attempts are successful for these teams.

```{r}
#| fig-cap: "The bar chart shows the count of shot attempts in each zone and whether the shot attempt was successful or not."
barplot4 = events_after_faceoff2 |>
  filter(eventOwnerTeamId == 12 | eventOwnerTeamId == 16 | eventOwnerTeamId == 5, !is.na(zoneCode)) |> 
  mutate(teams = case_when(eventOwnerTeamId == 12 ~ "CAR",
                           eventOwnerTeamId == 16 ~ "CHI",
                           eventOwnerTeamId == 5 ~ "PIT")) |>
  ggplot(aes(x = zoneCode, fill = as.factor(is_shot_atmpt))) +
  geom_bar() +
  facet_wrap(~ teams) +
  labs(fill = "Is Shot Attempt", title = "Shot Attempts between three NHL Teams") +
  theme(legend.position = "bottom")
barplot4
```

The following looks at a 2D kernel density estimation map of the offensive zone in hockey and where the future shot_attempts took place.

```{r}
future_shot_data = events_model |>
  select(xCoord, yCoord, future_shot) |>
  filter(xCoord >= 0) 

ggplot(data = future_shot_data, aes(x = xCoord, y = yCoord)) +
  geom_point(size = 1, alpha = 0.1) +
  geom_density_2d(lwd = 1.5) +
  coord_fixed()
```

Describe the data you’re using in detail, where you accessed it, along with relevant exploratory data analysis (EDA). You should also include descriptions of any relevant data pre-processing steps (e.g., whether you consider specific observations, create any meaningful features, etc.---but don't mention minor steps like column type conversion, filtering out unnecessary rows)

## Methods

Describe the modeling techniques you chose, their assumptions, justifications for why they are appropriate for the problem, and how you're comparing/evaluating the different methods.

### High Leverage Performance

The leverage model combines goal value and goal likelihood. Goal value is measured through win probability, or how valuable a goal is given the current game state. Goal likelihood is measured through the predicted xG generated within five seconds of a faceoff, or how much opportunity from each team’s perspective exists in a game state before the faceoff occurs. Therefore, the leverage model is not conditional on a faceoff win or loss. Instead, the variable represents the value of having the faceoff occur, regardless of the outcome.

#### Modeling Goal Value (Win Probability)

The first component of the leverage model, goal value, estimates win probability from each team's perspective given the current game state right before the faceoff. Predictors included goal differential, seconds remaining, zone, situation code, and a binary overtime status variable. (Link EDA plot) Exploratory analysis revealed the time remaining depends on the goal differential, motivating an interaction between the two variables. For example, the figure shows that the win probability for a team leading by one goal remains relatively stable, then spikes in the later stages of the game. Interactions between other features such as situation code and zone were considered but ultimately were not significant.

From there, the following three models were evaluated using five-fold cross-validation with folds assigned by game to maintain the mirrored faceoff data structure.

Logistic regression with linear time effects Logistic regression with natural spline time effects A generalized additive model (GAM) with score-specific smooths for time remaining

Model performance was evaluated using AUC, Brier score, and calibration. As expected, the nonlinear approaches outperformed the logistic regression (AUC = 0.817 vs. 0.812; Brier = 0.1717 vs. 0.1739). The spline model and GAM both adequately captured the relationship between goal differential and seconds remaining, producing virtually identical predictive performance and calibration.

Ultimately, the GAM was selected because, rather than specifying the spline degrees of freedom, the model estimates the shape of the nonlinear relationship from the data through penalized smoothing.

Three hypothetical game states from the win probability model are evaluated from (1) the current game state, (2) the current game state if the team scores a goal, and (3) the current game state if the team surrenders a goal. Calculating the change in win probability estimates the value of scoring a goal.

#### Modeling Goal Likelihood (Predicted xG)

Goal likelihood estimates the immediate scoring opportunity based on the current game state right before the faceoff. The goal of the leverage model measures the importance of the faceoff itself, so only pre-faceoff features were incorporated into the goal likelihood model, with one exception to account for the subsequent window of play. Therefore, the goal likelihood model is unconditional on a faceoff win or loss.

Just 621 goals were scored within five seconds of 370,890 faceoff events (0.0017%), making goals a highly variable response variable for the model. Expected goals (xG) served as a much more stable response in estimating immediate scoring opportunities following a faceoff. The distribution of xG within five seconds of a faceoff is heavily skewed, continuous, positive, and has a large spike of exact zeros (91.8% of faceoffs produced 0 xG). A Gaussian GAM was evaluated because xG is a continuous response variable. However, the Gaussian GAM predicted negative xG values for 16.6% of the faceoffs. Therefore, a Tweedie regression was a better distributional assumption for xG because of nonnegativity and a large mass at zero.

With the goal likelihood model focused on pre-faceoff variables, the same set of variables was selected as the goal value model. Once again, interactions between the variables were not significant. One additional feature added to the goal likelihood model was the available continuous play window for immediate scoring opportunity. Because the continuous window of play for most faceoffs was close to five seconds, the log of the continuous time window was used to compress larger values and increase resolution among shorter windows, resulting in a smoother relationship with scoring opportunity.

From there, the following three models were evaluated using five-fold cross-validation with folds assigned by game to maintain the mirrored faceoff data structure.

- Tweedie regression with linear time and available window effects

- Tweedie generalized additive model (GAM) with score-specific smooths for time remaining and available window

- Gaussian generalized additive model (GAM) with score-specific smooths for time remaining and available window

| Model        | RMSE   | MAE     |
|--------------|--------|---------|
| Tweedie GAM  | 0.0173 | 0.00386 |
| Gaussian GAM | 0.0173 | 0.00390 |
| Tweedie GLM  | 0.0174 | 0.00402 |

: Table comparing goal likelihood models

While the Gaussian and Tweedie GAMs produced nearly identical RMSE values, the Gaussian model generated impossible negative xG predictions. The Tweedie GAM achieved the lowest prediction error while avoiding negative xG values. The Tweedie GLM performed slightly worse, supporting the use of nonlinear smooth terms.

From there, offensive and defensive leverage components are calculated as:

- $\text{L\_Offense} = \Delta\text{WP\_GoalFor} \times \text{E(xGFor5)}$

- $\text{L\_Defense} = \Delta\text{WP\_GoalAgainst} \times \text{E(xGAgainst5)}$

- $\text{L} = \text{L\_Offense} + \text{L\_Defense}$

These two components are added to produce a leverage score, which is normalized to a 0-1 scale for interpretability.

Lastly, the normalized leverage scores were bucketed into three leverage groups: high, medium, and low. Jenks natural breaks clustering, a one-dimensional method, was selected to minimize within-group variance and maximize between-group variance. Three leverage groups were picked to balance interpretability with meaningful separation between game states.

#### Modeling Faceoff Win Value

The faceoff win value model incorporates the outcome of the faceoff to estimate the expected benefit of winning the draw. In other words, the faceoff win value model is contingent on a faceoff win or loss. Faceoff win value is calculated as the predicted change in net xG between winning and losing the same faceoff.

(Link back to EDA plot) Exploratory analysis showed that winning a faceoff in the offensive zone will produce a much higher xG value on average within five seconds of a faceoff compared to the neutral zone and defensive zone. Therefore, an interaction between winning the faceoff and zone code was added. Intuitively, this makes sense because there’s less distance from the attacking goal with a faceoff stationed in the offensive zone, especially if the winning team starts a clean possession.

Like the goal likelihood model, a Tweedie GAM evaluated xG For and xG Against within a five-second window to estimate the value of winning that faceoff. Each model included faceoff outcome (binary win/loss variable) interacted with zone code, along with goal differential, score-specific smooths (interaction between goal differential and seconds remaining), overtime variable, and the log-transformed available window.

Holding the current game state variables constant, two hypothetical scenarios were created by setting the outcome to either a win or a loss. The offensive and defensive values were measured as the difference in xG between these scenarios. An overall faceoff win value is computed by weighing both the offensive and defensive sides of the puck given the impact of a goal for or against in the current game state. Offensive and defensive values are combined to create a net xG faceoff win value.

### Quantifying Faceoff Value

#### Modeling Future Shot Probability

A simple logistic regression was first tried due to the binary variable of future unblocked shot attempt. However, when coordinates were looked at as a predictor, we decided to transition to a generalized additive model for smoothing. Due to the large sample size, the model was estimated using the `bam()` function from the **mgcv** package, which fits a GAM more efficiently. Thus, we modeled the probability of a future unblocked shot $(Y_i = 1)$ using a generalized additive model with a binomial error distribution and logit link function. The model was formulated as follows. $$\begin{aligned}
\log\left(\frac{P(\text{future\_shot}_i = 1)}{1 - P(\text{future\_shot}_i = 1)}\right) = \beta_0 &+ \beta_1(\text{eventTypeDescKey}_i) + s(\text{xCoord}_i, \text{yCoord}_i) + s(\text{distance}_i) \\&+ s(\text{secondsElapsedInGame}_i) + \beta_2(\text{strengthState}_i \times \text{scoreState}_i) \vphantom{\frac{1}{2}} \\&+ \beta_3(\text{isEmptyNetFor}_i) + \beta_4(\text{isEmptyNetAgainst}_i) + s(\text{angle}_i) + \varepsilon_i \vphantom{\frac{1}{2}}
\end{aligned}$$

The nonlinear effects of event location, distance and angle from the net, and seconds elapsed were modeled using thin plate regression splines with the basis dimensions (k) capped to avoid overfitting (k = 30 for coordinates, k = 20 for distance, and k = 15 for angle). The model also used fast Restricted Maximum Likelihood to cut down on processing time.

The data was split into a 75% training and a separate 25% testing. A single hold-out split was used instead of a k-fold cross-validation because of the long computational time associated with fitting large additive models. The model performance was then evaluated to check discrimination (ROC/AUC) and calibration. We checked calibration since we are trying to predict a future shot attempt following a sequence of events we need the calibration of the model to be precise.

## Results

Describe your results. This can include tables and plots showing your results, as well as text describing how your models worked and the appropriate interpretations of the relevant output. (Note: Don’t just write out the textbook interpretations of all model coefficients. Instead, interpret the output that is relevant for your question of interest that is framed in the introduction)

### High Leverage Performance

### Quantifying Faceoff Value

The future shot probability model resulted in high accuracy at predicting when a future shot after a faceoff will occur. The following plot is a calibration plot of the model showing how the model accurately predicts on the low and high end of shot probabilities, but is missing predictions in the middle range.

```{r}
shot_results = train_data |>
  select(future_shot, eventTypeDescKey, xCoord, yCoord,distance, scoreState, strengthState,
         secondsElapsedInGame, angle, isEmptyNetFor, isEmptyNetAgainst) |>
  drop_na() |>
  mutate(shot_prob = predict(nhl.mod, type = "response"),
         pred_decile2 = ntile(shot_prob, 10))

#checks each level of predictions withe the actual results
shot_calibration_check =  shot_results |>
  group_by(pred_decile2) |>
  summarize(
    predicted = mean(shot_prob),
    actual = mean(future_shot),
    n = n(),
    .groups = "drop")
#shot_calibration_check
ggplot(shot_calibration_check,
       aes(x = predicted, y = actual)) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed") +
  geom_line(color = "blue") +
  geom_point(shape = 15, size = 3, color = "blue", alpha = 0.5) +
  labs(x = "Predicted Probability", y = "Observed Probability",
       title = "Calibration Plot") +
  coord_equal() +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5,  size = 15))
```

The following plot is a ROC plot showing how the model is also quite accurate. The model had an AUC of 0.946, indicating good discrimination between predicting when a future unblocked shot attempt would occur and when it would not occur.

```{r}
roc_obj2 = roc(
  response = shot_results$future_shot,
  predictor = shot_results$shot_prob,
  quiet = TRUE)

auc(roc_obj2)

shot_roc = tibble(threshold = c(roc_obj2$thresholds),
                  specificity = roc_obj2$specificities,
                  sensitivity = roc_obj2$sensitivities)

shot_roc |> 
  ggplot(aes(x = 1 - specificity , y = sensitivity)) + #1-specificity (false pos. rate)
  geom_path() +
  geom_abline(slope = 1, intercept = 0, 
              linetype = "dashed") + 
  theme_bw() + 
  labs(title = "ROC Plot of Predicted Future Shot Attempts", y = "True Positive Rate", x = "False Positive Rate") +
  theme(plot.title = element_text(hjust = 0.5, size = 15))
```

## Discussion

Give your conclusions and summarize what you have learned with regards to your question of interest. Are there any limitations with the approaches you used? What do you think are the next steps to follow-up your project?

### High Leverage Performance

### Quantifying Faceoff Value

Ultimately we are able to determine if events directly following a faceoff will result in a future unblocked shot attempt. However, these probabilities are either on the very low end (\< 0.2) or on the higher end (\>0.60) most likely due to the use if o binary predictor.

## Appendix
### References
### Contact Information
